Graphical Tests for Power Comparison of Competing Designs

### 2012-03-30 intro.tex
Merge branch 'master' of github.com:heike/comparing-designs
Conflicts:
	setup.tex

The previous work \citet{buja:2009,wickham:2010} describes the use of
lineups for evaluating discoveries made in exploratory analysis. Here
we describe the use of lineups for evalating plot design for
communicating findings visually. Different lineups are created, using
the same data, same null plots and positions within the lineup, but
different plot designs. Comparing how long and how accurately the
viewers take to report the feature of interest will assess which
design is better for the task. In addition, this approach is ideally
coupled with simulation studies, where data is constructed to match
and expand on the features of interest in the real data. For example,
in a simple situation, where the discovered feature of interest is a
difference between the centers of two distributions, simulated data
would be constructed roughly matching the distribution of the real
data that varies the distance between centers, and also spread and
sample size. Lineups of these simulated data (embedded with simulated
null data plots) are generated and evaluated along with the lineups of
real data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a guage for
the signal strength of the pattern in the real data. It provides a way
to objectively evaluate plot design.

@@

Lineups provide a way to quantify the significance of the finding
\citet{majumder:2011}. 
%
The significance depends on the size ($m$) of
the lineup ($m=20$ in Figure \ref{lineup}) and the number of
independent observers ($n$). Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a Multinomial Distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$.

The power of a test is the probability to reject the null hypothesis. In the framework of lineups, power is the probability to identify the data plot, i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.

@@

\paragraph{Signal Strength}
To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength, and it is analogous to 1-$p$-value of a classical hypothesis test. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

>>>>

The previous work \citet{buja:2009,wickham:2010} describes the use of
lineups for evaluating discoveries made in exploratory analysis. Here
we describe the use of lineups for evalating plot design for
communicating findings visually. Different lineups are created, using
the same data, same null plots and positions within the lineup, but
different plot designs. Comparing how long and how accurately the
viewers take to report the feature of interest will assess which
design is better for the task. To investigate the benefit of lineups
for plot design evaluation we have conducted an experiment that also
contains simulated data, where the data is constructed to match and
expand on the features of interest in the real data. For example, in a
simple situation, where the discovered feature of interest is a
difference between the centers of two distributions, simulated data
would be constructed roughly matching the distribution of the real
data that varies the distance between centers, and also spread and
sample size. Lineups of these simulated data (embedded with simulated
null data plots) are generated and evaluated along with the lineups of
real data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a gauge for
the signal strength of the pattern in the real data.

@@

Lineups provide a way to statistically quantify the significance of the finding
\citet{majumder:2011}. Think of the linup as a set of $m$ plots, one of which is the real data. For one lineup and one observer, the significance level, $\alpha$ is $1/m$. This is the probability that the observer picks the plot of the real data by chance. It is the probability of picking the real data plot, when it is really *not* different from the other null plots, that is a Type I error. (In Figure \ref{lineup}, $m=20$.) The power of a test is the probability that the real data plot is identified when it really is different from the null plots. Power is usually difficult to calculate because it is requires specific use of alternative hypothesis to calculate the probability.  Work in \citet{niladri:2012} addresses this to some extent with measures on the quality of a lineup, how different numerically, as best it can be calculated, the real data plot is from the null plots. 
%
When there are multiple observers the significance level also depends on the number of independent observers ($n$). Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a multinomial distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$.

*** Under the null that the real is no different from the null, isn't $\pi_i=1/m$? Fixing this then I think the next paragraph on signal strength flows right from this one. It is when we are calculating signal strength that we start counting the number of observers who pick it.

%The power of a test is the probability to reject the null hypothesis, when it is not true. In the framework of lineups, power is, effectively, the probability that the data plot is identified. , i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.

@@

\paragraph{Signal Strength}
To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength \citet{majumder:2012}, and it is analogous to 1-$p$-value of a classical hypothesis test. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.
